Nonlinear Tidal Dissipation in White Dwarfs

### 2017-12-11 proposal/proposal.tex
outline
try proposal

    \documentclass[12pt,
        usenames, % allows access to some tikz colors
        dvipsnames % more colors: https://en.wikibooks.org/wiki/LaTeX/Colors
    ]{article}
    \usepackage{
        amsmath,
        amssymb,
        fancyhdr, % page styling
        lastpage, % footer fanciness
        hyperref, % various links
        setspace, % line spacing
        amsthm, % newtheorem and proof environment
        mathtools, % \Aboxed for boxing inside aligns, among others
        float, % Allow [H] figure env alignment
        enumerate, % Allow custom enumerate numbering
        graphicx, % allow includegraphics with more filetypes
        wasysym, % \smiley!
        upgreek, % \upmu for \mum macro
        listings, % writing TrueType fonts and including code prettily
        tikz, % drawing things
        times,
        booktabs % \bottomrule instead of hline apparently
    }
    \usepackage[margin=0.5in, top=1in, bottom=1in]{geometry} % page geometry
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[font=scriptsize]{subcaption} % subfigures

@@

    \everymath{\displaystyle} % biggify limits of inline sums and integrals
    \tikzstyle{circ} % usage: \node[circ, placement] (label) {text};
        = [draw, circle, fill=white, node distance=3cm, minimum height=2em]
    \definecolor{commentgreen}{rgb}{0,0.6,0}
    \lstset{
        basicstyle=\ttfamily\footnotesize,
        frame=single,
        numbers=left,
        showstringspaces=false,
        keywordstyle=\color{blue},
        stringstyle=\color{purple},
        commentstyle=\color{commentgreen},
        morecomment=[l][\color{magenta}]{\#}
    }

@@

\doublespacing

@@

@@

In the last decade, astrophysical systems with one white dwarf and another
body (e.g.\ a black hole or another white dwarf), referred to collectively as WD
binaries, have become increasingly important for several topics in astrophysics.
Mergers of WD-WD binaries create a variety of astrophysical systems, the most
important of which are type Ia supernovae, the standard candle of observational
astronomy. WD-BH (black hole) systems are expected to produce detectable flares
that can be observed to yield constraints on currently unknown physics
(https://arxiv.org/pdf/1701.08162.pdf).

\emph{Tidal interactions}, the effect of an uneven gravitational field produced
by one body on the other, are the key component in such systems. In WD binaries,
tidal interactions are expected to excite internal gravity waves in the WD\@.
These then propagate towards the surface and dissipate by wave breaking, as ocean
waves do on a beach, among other mechanisms.
Prior work shows that this wave breaking can play a dominant role in determining
the internal structure of the WD\@. Nonetheless, previous treatments of WD
binaries adopt incomplete, semi-analytical treatments to estimate the location
and strength of wave-breaking induced dissipation. For such fluid systems,
numerical simulation is the only way to obtain quantitatively accurate results.
% (Fuller 2012d)

\textbf{We propose to study nonlinear wave breaking of internal gravity waves in
WDs using numerical hydrodynamic codes.} There are already a wide assortment of
suitable hydrodynamical codes
% (Dedalus\footnote{
% http://iopscience.iop.org/article/10.1088/0004-%
% 637X/796/1/17}, Athena\footnote{e.g.\
% http://iopscience.iop.org/article/10.1088/0004-%
% 637X/741/1/57})
capable of solving the fully nonlinear hydrodynamical equations, we expect to be
able to apply such codes with slight modifications to solve the proposed
problem. Should existing codes prove inapplicable, the applicant is well
equipped to develop suitable software from scratch (\autoref{s:plan}). We aim
to extend our findings to observational signatures for a variety of compositions
and uncertain material properties.

The results of our study as GW astronomy continues to improve; inaccuracies are
especially important to mitigate with the high sensitivity requirements
involved. Conversely, a gravitational wave detection could yield significant
information on WD dynamics that, thanks to their ubiquity, could have
ramifications throughout astronomy. Equally importantly, our approach would be
highly transferable to other problems such as tidal heating during exoplanet
formation, another area of crucial and open research.

A tentative timeline for research follows:
\begin{itemize}
    \item 2--4 months: Set up hydrodynamic code, calibrate against simplified,
        well-understood fiducial models.
    \item 4--10 months: Apply hydrodynamic code to systems found in literature.
    \item 10--14 months: Generate observational counterparts to predicted
        phenomena.
    \item 14--24 months: Explore related interesting astrophysical systems, seek
        analytical descriptions of any discovered phenomena.
\end{itemize}

\section{Plan of Study}\label{s:plan}

Cornell University is the ideal institution to pursue the proposed research.
Professor Dong Lai, the applicant's adviser, has authored numerous papers on WD
tidal dissipation and is a foremost expert in the subject. Other faculty and
researchers at Cornell are equally well-suited for the project. Notably,
Professor Saul Teukolsky is a leading researcher in numerical techniques in
astrophysics for the past five decades.

The applicant earned two B.S.\ from the California Institute of Technology in
Physics and Computer science and is well equipped for the proposed problem with
teaching proficiency in differential equations and completed coursework in
computational physics, astrophysical fluid dynamics, GPU programming and others.
Future relevant coursework includes astrophysical processes, celestial
mechanics and physics of compact objects, all of which are scheduled to be
completed within six months of recipt of the fellowship funding. These courses
will enable both a complete treatment of relevant physics and accurate, fast
programming of numerical codes.

The proposed study is integral to the long-term career goals of the applicant,
a professorship in astrophysics at a research institution. Proficiency in
numerical techniques and software development is extremely important in modern
astrophysical studies. White dwarfs both are important to observational
astronomy and involve many astrophysical processes, so such a project would be
relevant to many members of the academic community and open avenues for
collaboration.

% \begin{itemize}
%     \item Lots of WKB treatments
%     % TODO radiative diffusion vs turbulent dissipation?
%     %   - if in stars can power energic outbursts, same for WD?
%     % Planet formation, WD inspirals + waveforms (important b/c Fuller et al P4)
%     %   - heating should have visible signatures in flares
%     \item

%     \item \url{http://journals.ametsoc.org/doi/pdf/10.1175/1520-%
%         0469\%281983\%29040\%3C2497\%3ATIOGWB\%3E2.0.CO\%3B2} Wave breaking is
%         an essential part of circulation % holten et al 1983

%     \item \url{https://arxiv.org/pdf/1211.0624.pdf}, % fuller 2012, vick 2016
%         \url{https://arxiv.org/abs/1612.07316} Show necessary to get heating as
%         function of depth

%     \item \url{http://journals.ametsoc.org/doi/abs/10.1175/JAS-D-11%
%         -097.1} has already studied to some extent in atmospheres, weakly
%         nonlinear regime
%         % sutherland 2011

%     \item Tidal heating in planet formation too!
% \end{itemize}

>>>>

    \documentclass[11pt,
        usenames, % allows access to some tikz colors
        dvipsnames % more colors: https://en.wikibooks.org/wiki/LaTeX/Colors
    ]{article}
    \usepackage[numbers]{natbib}
    \usepackage{
        amsmath,
        amssymb,
        fouriernc,
        fancyhdr, % page styling
        lastpage, % footer fanciness
        hyperref, % various links
        setspace, % line spacing
        amsthm, % newtheorem and proof environment
        mathtools, % \Aboxed for boxing inside aligns, among others
        float, % Allow [H] figure env alignment
        enumerate, % Allow custom enumerate numbering
        graphicx, % allow includegraphics with more filetypes
        wasysym, % \smiley!
        upgreek, % \upmu for \mum macro
        listings, % writing TrueType fonts and including code prettily
        tikz, % drawing things
        times,
        booktabs % \bottomrule instead of hline apparently
    }
    \usepackage[margin=0.8in, top=1in, bottom=1in]{geometry} % page geometry
    \usepackage[
        labelfont=bf, % caption names are labeled in bold
        font=scriptsize % smaller font for captions
    ]{caption}
    \usepackage[font=scriptsize]{subcaption} % subfigures

@@

    \tikzstyle{circ} % usage: \node[circ, placement] (label) {text};
        = [draw, circle, fill=white, node distance=3cm, minimum height=2em]
    \definecolor{commentgreen}{rgb}{0,0.6,0}
    \lstset{
        basicstyle=\ttfamily\footnotesize,
        frame=single,
        numbers=left,
        showstringspaces=false,
        keywordstyle=\color{blue},
        stringstyle=\color{purple},
        commentstyle=\color{commentgreen},
        morecomment=[l][\color{magenta}]{\#}
    }

@@

\onehalfspacing

@@

\title{Nonlinear Tidal Dissipation in White Dwarfs}
\author{Yubo Su}
\date{}

\maketitle

\section{Introduction: White Dwarf Binaries}

White dwarfs (WDs) are luminous, long-lasting products of late-stage stellar
evolution that are held up by electron degeneracy pressure, a quantum mechanical
effect arising from subjecting electrons to immense pressures. They are some of
the densest objects in the known universe, behind only neutron stars (NS) and
black holes (BH), fitting a solar mass into an Earth-sized sphere. Furthermore,
they exhibit various chemical compositions, each of which is subject to
different physics, and are found in a variety of interesting systems. As such,
WDs are a unique and diverse window into matter under extreme conditions.

WDs are commonly found in binary orbits with companion objects, in which the two
orbit their center of mass under their mutual gravitational attraction. The
companion object ranges from another WD to a supermassive BH (SMBH),
hypothesized to be at the center of galaxies and have mass $10^5$--$10^6$ solar
masses. All of these binary systems are very important to astrophysics. WD-WD
binaries are most important for being thought to generate \emph{Type Ia
supernovae} in which one of the two WDs reaches a critical mass through
accreting surrounding matter and collapses under its own gravity, producing a
shock wave that tears the WD apart and releases immense amounts of energy. Type
Ia supernovae are unique in astrophysics for their near-constancy throughout
space and time, consistently occuring at the WD critical mass, and have been
used to discover numerous cosmological phenomena such as dark energy.

WD-BH systems are also interesting subjects of study. Studies indicate that
as WDs orbit the extreme gravity surrounding BHs, they will produce observable
flares induced by gravitational tidal forces\cite{flares}. A WD orbiting a SMBH
would produce gravitational waves, waves carried by distortions in space-time
and generated according to Einstein's general relativity by accelerating massive
objects, that are expected to be detected by the space-based \emph{Laser
Interferometer Space Antenna (LISA)} when deployed\cite{lisa}. Gravitational
wave astronomy is an increasingly exciting field as the \emph{Laser
Interferometer Gravitational-Wave Observatory (LIGO)} continues to make
detections since its first in late 2015. As gravitational wave astronomy relies
on accurate predictions of the expected signals, it is important to build as
accurate models as possible before observation runs begin.

\subsection{Tidal Dissipation}

The excitation of \emph{internal gravity waves} in the WD by the tidal forces of
the companion is an effect seen in all of the aforementioned systems. Internal
gravity waves, not to be confused with the gravitational waves discussed above,
are internal displacements in the WD that oscillate and propagate due to a
restoring buoyancy force. As these waves propagate outward from where they are
excited, they are expected to grow in amplitude until they break, as do ocean
waves on a shore, and deposit both energy and angular momentum inside the WD\@.

Previous work predicts that this dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
are thus a significant contribution to the WD energy budget\cite{fullerII}. The
exact radial dissipation profile is of enormous interest since it is both
sensitive to WD properties and can produce drastically different observable
outcomes. One proposed outcome is a \emph{tidal nova}, in which heating in the
WD's degenerate hydrogen layer is significant enough to trigger runaway nuclear
fusion and a significant, observable explosion\cite{tidal_novae}. Understanding
whether such phenomena occur requires understanding how energy is distributed
internally inside a WD\@.

Internal gravity wave breaking is a nonlinear hydrodynamic phenomenon. Such
phenomena are known to require numerical simulation to study. It is therefore
paramount to begin numerical study to build dissipation models inside WDs to
characterise what phenomena can be observed in which WD models and
characteristics.

@@

\textbf{We propose to study tidal dissipation via nonlinear gravity wave
breaking in white dwarfs via numerical simulation.}

Our research will consist of computing for various WD models and compositions
the energy and angular momentum dissipation profiles inside WDs. Once such
profiles are obtained, we intend to add these profiles to existing stellar
evolution codes to study the dynamical effects of tidal dissipation. We will
attempt to find a compact or even analytical representation of our numerical
work to greatly simplify and accelerate such integration. Finally, we will make
observational predictions with our results and compare our predictions with
existing observed WDs. Any software and results would be made in accordance with
best practice and publicly available for the entire community.

\subsection{Qualifications}

The proposing researcher (YS) and his adviser (DL) are uniquely qualified to
pursue such studies. YS has ample research experience in numerical simulation
and is continuing to pursue academic study in hydrodynamics, turbulence and
computation. He also completed a double Bachelors degree in physics and computer
science from the California Institute of Technology. Finally, he has worked in
the software industry for a year and continues to follow discussions of best
practice far ahead of those in academia and apply them to his own work.

DL is a co-author on much of the latest literature on WD binaries and is an
expert on many related subjects such as fluid dynamics in extreme matter and
studies of other compact object systems. Cornell University also houses Saul
Teukolsky's research group, one of the pre-eminent numerical relativity groups
in the world, which will be frequently consulted for numerical best practices.

In terms of software, YS proposes to only implement what is not currently
available in the community. Current literature indicates that the spectral
hydrodynamic code Dedalus is able to handle instability-driven turbulence with
high accuracy and speed, and our preliminary work seems to indicate that Dedalus
will be sufficiently accurate to simulate wave breaking
dissipation\cite{dedalus}. Modules for Experiments in Stellar Astrophysics
(MESA) is a proven stellar evolution code that will likely need only be extended
to incorporate the discovered dissipation models\cite{MESA}. Nevertheless, both
codes will require adaptation to be applied to the present problem (YS has
already contributed an improvement to Dedalus), a task for which YS is
well-equiped.

Should no existing software prove adequate, we are more than able and willing to
develop a new code and contribute it to the community. Notably, no
GPU-accelerated spectral hydrodynamic code has yet been released, which would be
of great value to researchers, and YS is well-versed in GPU programming.
Nevertheless, in the interest of the proposed science objectives codes will be
developed on an as-needed basis.

\subsection{Applications, Interest to the DoD}

A robust pipeline coupling nonlinear wave breaking to system evolution
would be of interest spanning all fields of physics. Barring BHs, all
astrophysical objects exhibit internal waves that can break as they propagate
towards the surface. Notable present interests in the astrophysical community
include tidal disruption, planet formation and stars. Even in planetary and
atmospheric sciences, wave breaking in atmospheres would share many common
techniques with the proposed research.

The proposed research is extremely relevant to Department of Defense (DoD)
fields of interest. WDs and its energetic phenomena we propose to study e.g.\
novae of all sorts are natural laboratories of extreme physics that have
direct consequences for understanding of plasmas and other technologies integral
to propulsion in aeronautics. A theory of nonlinear wave breaking would have
far-reaching consequences in turbulence study and atmospheric sciences. Finally,
the observation of our predictions could shed light on atmospheric optics,
whether as a use for calibration or understanding contamination effects.

\bibliographystyle{plainnat}
\renewcommand{\bibname}{References}
{\scriptsize \bibliography{proposal}}
